---
title: "Comparing Risk and Return Across Equity ETFs, Gold, and Bitcoin (2016-2025)"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## 1. Motivation

Every long-term investor faces two practical questions: how much diversification
benefit do "alternative" assets like gold and Bitcoin actually provide alongside
standard equity holdings, and have higher-risk assets historically been
compensated with proportionally higher risk-adjusted returns? This project
compares six investment assets over the past ten years
(Jan 1, 2016 - Dec 31, 2025): VOO (S&P 500), QQQ (NASDAQ-100),
IWM (Russell 2000), VTI (total U.S. market), GLD (gold), and BTC-USD (Bitcoin).

## 2. Setup

```{r setup}
library(tidyverse)
library(tidyquant)
library(scales)
library(knitr)

tickers    <- c("VOO", "QQQ", "IWM", "VTI", "GLD", "BTC-USD")
start_date <- as.Date("2016-01-01")
end_date   <- as.Date("2025-12-31")

theme_set(theme_minimal(base_size = 11))
```

## 3. Data Acquisition

The project uses two different types of data sources, as required by the rubric.

### 3.1 Source 1: Yahoo Finance API (via tidyquant)

```{r get-prices}
prices_raw <- tq_get(tickers,
                     get  = "stock.prices",
                     from = start_date,
                     to   = end_date)

prices <- prices_raw |>
  mutate(symbol = if_else(symbol == "BTC-USD", "BTC", symbol)) |>
  select(symbol, date, adjusted)

glimpse(prices)
```

### 3.2 Source 2: FRED CSV downloads

CPI (Consumer Price Index) is used to compute inflation-adjusted (real) returns, and the 3-month
Treasury rate is used as the risk-free rate in Sharpe ratio calculations.
Both are pulled directly from FRED's (Federal Reserve Economic Data) CSV endpoints.

```{r get-fred}
# Original data source: https://fred.stlouisfed.org/graph/fredgraph.csv?id=CPIAUCSL
cpi <- read_csv(
  "https://raw.githubusercontent.com/qingquan-li/Data607/refs/heads/main/final-project/data/CPIAUCSL.csv",
  show_col_types = FALSE
) |>
  rename(date = 1, cpi = 2) |>
  mutate(date = as.Date(date)) |>
  filter(date >= start_date, date <= end_date)

# Original data source: https://fred.stlouisfed.org/graph/fredgraph.csv?id=DGS3MO
rf <- read_csv(
  "https://raw.githubusercontent.com/qingquan-li/Data607/refs/heads/main/final-project/data/DGS3MO.csv",
  show_col_types = FALSE,
  na = c(".", "")
) |>
  rename(date = 1, rf_pct = 2) |>
  mutate(date   = as.Date(date),
         rf_pct = as.numeric(rf_pct)) |>
  filter(date >= start_date, date <= end_date) |>
  drop_na()

head(cpi)
head(rf)
```

## 4. Data Transformation

This step satisfies the "at least one data transformation" requirement:
prices are transformed to daily simple returns and pivoted from long to wide
format so each asset becomes its own column for cross-asset analysis.

```{r transform}
# Daily simple returns (long format)
returns_long <- prices |>
  group_by(symbol) |>
  arrange(date) |>
  mutate(return = adjusted / lag(adjusted) - 1) |>
  filter(!is.na(return)) |>
  ungroup()

# Pivot to wide format; keep only days where ALL assets traded
# (excludes BTC weekend-only days for fair cross-asset comparison)
returns_wide <- returns_long |>
  select(symbol, date, return) |>
  pivot_wider(names_from = symbol, values_from = return) |>
  drop_na()

cat("Trading days in aligned panel:", nrow(returns_wide), "\n")
cat("Date range:",
    format(min(returns_wide$date)), "to",
    format(max(returns_wide$date)), "\n")
```

## 5. Exploratory Analysis (Describing the Data)

### 5.1 Growth of $1 invested at the start of 2016

```{r growth-chart, fig.width=9, fig.height=5}
growth <- returns_wide |>
  pivot_longer(-date, names_to = "symbol", values_to = "return") |>
  group_by(symbol) |>
  arrange(date) |>
  mutate(value = cumprod(1 + return)) |>
  ungroup()

ggplot(growth, aes(date, value, color = symbol)) +
  geom_line(linewidth = 0.6) +
  scale_y_log10(labels = dollar_format()) +
  labs(title = "Growth of $1 invested at the start of 2016 (log scale)",
       x = NULL, y = "Portfolio value", color = "Asset")
```

### 5.2 Daily return distributions

```{r return-distributions, fig.width=9, fig.height=5}
returns_wide |>
  pivot_longer(-date, names_to = "symbol", values_to = "return") |>
  ggplot(aes(return, fill = symbol)) +
  geom_histogram(bins = 80, alpha = 0.85) +
  facet_wrap(~ symbol, scales = "free") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "Daily return distributions, 2016-2025",
       x = "Daily return", y = "Count") +
  theme(legend.position = "none")
```

BTC has visibly fatter tails than the equity ETFs (large gains and losses
of more than ten percent in a single day), while GLD looks closest to a
narrow, symmetric distribution.

## 6. Statistical Analysis (Supporting Conclusions)

### 6.1 Summary statistics by asset

```{r summary-stats}
# Helper: maximum drawdown from a vector of simple returns
max_drawdown <- function(r) {
  v <- cumprod(1 + r)
  abs(min((v - cummax(v)) / cummax(v)))
}

avg_rf <- mean(rf$rf_pct, na.rm = TRUE) / 100  # average annualized risk-free rate

summary_stats <- returns_wide |>
  pivot_longer(-date, names_to = "symbol", values_to = "return") |>
  group_by(symbol) |>
  summarize(
    obs               = n(),
    annualized_return = mean(return) * 252,
    annualized_vol    = sd(return)   * sqrt(252),
    sharpe            = (annualized_return - avg_rf) / annualized_vol,
    max_drawdown      = max_drawdown(return),
    .groups = "drop"
  ) |>
  arrange(desc(sharpe))

summary_stats |>
  mutate(across(annualized_return:max_drawdown, ~ round(.x, 3))) |>
  kable(caption = "Annualized return, volatility, Sharpe ratio, and max drawdown by asset (2016-2025)")
```

### 6.2 Correlation matrix (diversification check)

```{r correlation-heatmap, fig.width=7, fig.height=6}
cor_matrix <- returns_wide |>
  select(-date) |>
  cor()

cor_long <- as.data.frame(cor_matrix) |>
  rownames_to_column("asset1") |>
  pivot_longer(-asset1, names_to = "asset2", values_to = "correlation")

ggplot(cor_long, aes(asset1, asset2, fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", correlation)),
            color = "white", size = 4) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Daily return correlations, 2016-2025",
       x = NULL, y = NULL, fill = "Correlation")
```

The three equity ETFs (VOO, QQQ, VTI, IWM) are tightly correlated with each
other (typically above 0.85), GLD is roughly uncorrelated with equities,
and BTC shows a weak positive correlation with equities.

### 6.3 CAPM regression: beta vs the S&P 500

This regresses each asset's daily excess return on VOO's daily excess return,
producing a CAPM-style beta and alpha for each asset.

```{r capm}
returns_excess <- returns_wide |>
  left_join(rf |> select(date, rf_pct), by = "date") |>
  fill(rf_pct, .direction = "down") |>
  mutate(rf_daily = (rf_pct / 100) / 252) |>
  mutate(across(c(VOO, QQQ, IWM, VTI, GLD, BTC),
                ~ .x - rf_daily,
                .names = "ex_{.col}"))

assets_to_test <- c("QQQ", "IWM", "VTI", "GLD", "BTC")

capm_results <- map_dfr(assets_to_test, function(asset) {
  fit <- lm(reformulate("ex_VOO", paste0("ex_", asset)),
            data = returns_excess)
  s <- summary(fit)
  tibble(
    asset         = asset,
    alpha_daily   = coef(fit)[1],
    beta          = coef(fit)[2],
    r_squared     = s$r.squared,
    p_value_alpha = s$coefficients[1, 4]
  )
})

capm_results |>
  mutate(across(where(is.numeric), ~ round(.x, 4))) |>
  kable(caption = "CAPM regression of each asset's daily excess return on VOO")
```

Equity ETFs (QQQ, VTI, IWM) all have betas close to or above 1, as expected.
GLD's beta is near zero (gold barely moves with the stock market on a daily
basis), and BTC has a modest positive beta with very low R-squared, meaning
most of its variation is unexplained by stock-market moves.

## 7. Feature Beyond Class: Rolling 12-Month Correlations

This section uses a rolling 252-trading-day window to track how the
correlation between Bitcoin and equities (and gold and equities) has shifted
over the period - something a static correlation matrix cannot reveal.

```{r rolling-corr, fig.width=9, fig.height=5}
# Manual rolling correlation (no extra packages needed)
rolling_cor <- function(x, y, n = 252) {
  out <- rep(NA_real_, length(x))
  for (i in n:length(x)) {
    out[i] <- cor(x[(i - n + 1):i], y[(i - n + 1):i])
  }
  out
}

rolling_corr <- returns_wide |>
  arrange(date) |>
  mutate(
    `BTC vs VOO` = rolling_cor(BTC, VOO),
    `BTC vs QQQ` = rolling_cor(BTC, QQQ),
    `GLD vs VOO` = rolling_cor(GLD, VOO)
  ) |>
  select(date, `BTC vs VOO`, `BTC vs QQQ`, `GLD vs VOO`) |>
  drop_na() |>
  pivot_longer(-date, names_to = "pair", values_to = "correlation")

ggplot(rolling_corr, aes(date, correlation, color = pair)) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Rolling 12-month (252-day) correlations with equities",
       x = NULL, y = "Correlation", color = NULL)
```

The rolling chart shows that BTC's correlation with stocks was near zero for
most of 2016-2019, then climbed sharply in 2020-2022 during the era of
synchronized macro liquidity, before easing again. Gold's correlation with
equities stayed near zero across the entire period - it is the more
consistent diversifier, while Bitcoin's diversification benefit has been
regime-dependent.

## 8. Inflation-Adjusted (Real) Returns

A quick sanity check using the second source (FRED CPI): converting nominal
cumulative returns to real returns.

```{r real-returns}
# Pull start and end CPI to compute total inflation over the period
cpi_start <- cpi |> arrange(date) |> slice(1)  |> pull(cpi)
cpi_end   <- cpi |> arrange(date) |> slice(n()) |> pull(cpi)
total_inflation <- cpi_end / cpi_start - 1

cat(sprintf("Total CPI inflation, %s to %s: %.1f%%\n",
            format(min(cpi$date)), format(max(cpi$date)),
            100 * total_inflation))

nominal_growth <- growth |>
  group_by(symbol) |>
  summarize(nominal_final = last(value), .groups = "drop")

real_returns <- nominal_growth |>
  mutate(real_final           = nominal_final / (1 + total_inflation),
         nominal_total_return = nominal_final - 1,
         real_total_return    = real_final    - 1) |>
  arrange(desc(real_total_return))

real_returns |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  kable(caption = "Nominal vs real (inflation-adjusted) total return, 2016-2025")
```

## 9. Conclusions

A few clear takeaways from the analysis:

1. **Tech concentration paid off in this window.** QQQ posted the highest
   annualized return of the equity ETFs, but it also had the largest
   drawdown among them. Concentration cuts both ways.
2. **Bitcoin had the highest absolute return but also the worst risk-adjusted
   profile by a wide margin.** Its annualized volatility and maximum drawdown
   dwarf those of every other asset, so its high Sharpe ratio is sensitive
   to the start and end dates of the window.
3. **Gold was the more reliable diversifier.** Its correlation with equities
   stayed near zero across the entire period, while BTC's correlation
   climbed materially after 2020, weakening its diversification value
   precisely when it would have been most useful.
4. **The three broad-market equity ETFs (VOO, VTI, IWM) are highly
   correlated with each other.** Holding all three is closer to a
   concentrated bet on U.S. equities than a diversified portfolio.
