---
title: "document-classification"
format:
  html: "default"
  pdf:
    number-sections: true
---

# Introduction

The goal is to take a set of e-mails that have already been hand-labeled
as "ham" (legitimate) or "spam" and learn a model that can label new,
unseen e-mails. The corpus used here is the public SpamAssassin corpus
(https://spamassassin.apache.org/old/publiccorpus/).

Approach:

1. Get the data: download the `easy_ham` and `spam` archives from the
   SpamAssassin public corpus and extract them locally.
2. Load and label each e-mail file with its class (`ham` or `spam`).
3. Preprocess the text: lowercase, strip numbers and punctuation,
   remove English stopwords, and stem the remaining tokens.
4. Build a document-term matrix (rows = e-mails, columns = tokens) and
   drop tokens that are too rare to be informative.
5. Split 75 / 25 into train and test sets so the model is evaluated on
   documents it has never seen.
6. Fit a Naive Bayes classifier. Naive Bayes treats each token as
   independent evidence for or against "spam" and combines them with
   Bayes' rule -- it is the textbook baseline for bag-of-words spam
   filtering because it is fast, simple, and surprisingly competitive.
7. Evaluate with a confusion matrix and accuracy / precision / recall,
   then peek at the most informative tokens as a sanity check.

# Solution

## Setup

```{r setup}
library(tm)
library(SnowballC)
library(e1071)
library(caret)

set.seed(42)
```

## Download and extract the corpus

```{r download}
data_dir <- "spamassassin"
dir.create(data_dir, showWarnings = FALSE)

urls <- list(
  ham  = "https://spamassassin.apache.org/old/publiccorpus/20030228_easy_ham.tar.bz2",
  spam = "https://spamassassin.apache.org/old/publiccorpus/20030228_spam.tar.bz2"
)

for (label in names(urls)) {
  archive <- file.path(data_dir, paste0(label, ".tar.bz2"))
  if (!file.exists(archive)) {
    download.file(urls[[label]], archive, mode = "wb", quiet = TRUE)
    untar(archive, exdir = data_dir)
  }
}
```

## Load e-mails

Each e-mail file is read as a single text string. We use `latin1`
encoding because spam often contains bytes that are not valid UTF-8,
and we want to read the file rather than crash on it.

```{r load}
ham_dir  <- file.path(data_dir, "easy_ham")
spam_dir <- file.path(data_dir, "spam")

read_email <- function(f) {
  paste(readLines(f, warn = FALSE, encoding = "latin1"), collapse = " ")
}

ham_files  <- list.files(ham_dir,  full.names = TRUE)
spam_files <- list.files(spam_dir, full.names = TRUE)

# Drop the index files that ship with the archive
ham_files  <- ham_files[!grepl("cmds$",  ham_files)]
spam_files <- spam_files[!grepl("cmds$", spam_files)]

emails <- c(vapply(ham_files,  read_email, character(1)),
            vapply(spam_files, read_email, character(1)))
labels <- factor(c(rep("ham",  length(ham_files)),
                   rep("spam", length(spam_files))))

cat("ham:", length(ham_files), " spam:", length(spam_files), "\n")
```

## Clean the text and build the document-term matrix

```{r dtm}
corpus <- VCorpus(VectorSource(emails))
corpus <- tm_map(corpus, content_transformer(tolower))
corpus <- tm_map(corpus, removeNumbers)
corpus <- tm_map(corpus, removePunctuation)
corpus <- tm_map(corpus, removeWords, stopwords("english"))
corpus <- tm_map(corpus, stemDocument)
corpus <- tm_map(corpus, stripWhitespace)

dtm <- DocumentTermMatrix(corpus)
dtm <- removeSparseTerms(dtm, 0.95)   # keep tokens in at least 5% of docs
dim(dtm)
```

## Convert counts to presence/absence and split

For a Y/N feature, Naive Bayes estimates `P(token present | class)`
directly, which is exactly what a bag-of-words spam filter needs.

```{r split}
mat <- ifelse(as.matrix(dtm) > 0, "Y", "N")
df  <- as.data.frame(mat, stringsAsFactors = FALSE)
df[] <- lapply(df, factor, levels = c("N", "Y"))

n         <- nrow(df)
train_idx <- sample(seq_len(n), size = floor(0.75 * n))

x_train <- df[train_idx, ]
x_test  <- df[-train_idx, ]
y_train <- labels[train_idx]
y_test  <- labels[-train_idx]
```

## Train and evaluate

```{r model}
model <- naiveBayes(x_train, y_train, laplace = 1)
pred  <- predict(model, x_test)

cm <- confusionMatrix(pred, y_test, positive = "spam")
cm$table
round(cm$overall["Accuracy"], 3)
round(cm$byClass[c("Sensitivity", "Specificity", "Precision", "F1")], 3)
```

## Most informative tokens

Sanity check -- which stemmed tokens have the strongest "spam" signal
under the trained model:

```{r top-tokens}
spam_prob <- sapply(model$tables, function(tbl) tbl["spam", "Y"])
ham_prob  <- sapply(model$tables, function(tbl) tbl["ham",  "Y"])
top <- sort(spam_prob - ham_prob, decreasing = TRUE)[1:15]
round(top, 3)
```

# Conclusion

A bag-of-words plus Naive Bayes pipeline produces a respectable spam
classifier in very little code: clean the text, count tokens, hold out
a test set, fit, evaluate. On the SpamAssassin `easy_ham` + `spam`
split this reaches accuracy in the high 90s, and the most informative
tokens line up with intuition (price words, click-bait, HTML tag
fragments).

Caveats worth flagging:

- Naive Bayes assumes tokens are independent given the class, which is
  false (e.g. "free" and "money" co-occur). The model still works
  because the wrong-but-cheap probabilities still rank documents in
  roughly the right order.
- `easy_ham` is, as the name implies, easy. The `hard_ham` split would
  drag accuracy down and give a more honest picture of real-world
  performance.
- Headers were left in as raw text. Splitting headers from the body and
  engineering header-specific features (sender domain, Received chain,
  Content-Type) would likely help a lot.
- Production filters layer URL reputation, sender authentication
  (SPF / DKIM / DMARC), and user feedback on top of a content model
  like this rather than relying on it alone.
