---
title: "week-2A-SQL-and-R"
format: html
---

# Approach

I will collect a small movie-ratings dataset by selecting six recent popular movies and asking at least five people to rate the movies they’ve seen on a 1–5 scale. I will store the results in a PostgreSQL database, using a simple schema (3 tables: users, movies, ratings with primary keys and foreign keys) and SQL scripts to create/populate the tables. Then I will load the ratings into R as a dataframe (either by querying the database from R or exporting to CSV first) and produce basic summaries such as average ratings by movie and counts of ratings per user/movie.

The main data challenge I anticipate is missing ratings, since not everyone will have seen every movie. I will handle this explicitly, for example, represent missing ratings as NULL in SQL, and decide whether to exclude missing values from averages or impute a reasonable placeholder depending on the metric.

# 6 recent popular movies

1.  Dune: Part Two (2024)
2.  Barbie (2023)
3.  Oppenheimer (2023)
4.  Killers of the Flower Moon (2023)
5.  The Batman (2022)
6.  Spider-Man: Across the Spider-Verse (2023)

# Setup PostgreSQL database tables/data

## Create database

```sql
CREATE DATABASE movie_ratings;

-- Connect to it
\c movie_ratings
```

## Create tables

```sql
CREATE TABLE people (
  person_id SERIAL PRIMARY KEY,
  name TEXT NOT NULL UNIQUE
);

CREATE TABLE movies (
  movie_id SERIAL PRIMARY KEY,
  title TEXT NOT NULL UNIQUE,
  release_year INT NOT NULL
);

CREATE TABLE ratings (
  rating_id SERIAL PRIMARY KEY,
  person_id INT NOT NULL REFERENCES people(person_id),
  movie_id  INT NOT NULL REFERENCES movies(movie_id),
  rating INT NULL CHECK (rating BETWEEN 1 AND 5),
  UNIQUE (person_id, movie_id)
);
```

## Insert people and movies

```sql
INSERT INTO people (name) VALUES
('Jacob'),
('Alice'),
('Ben'),
('Chloe'),
('David');

INSERT INTO movies (title, release_year) VALUES
('Dune: Part Two', 2024),
('Barbie', 2023),
('Oppenheimer', 2023),
('Killers of the Flower Moon', 2023),
('The Batman', 2022),
('Spider-Man: Across the Spider-Verse', 2023);
```

## Insert ratings with missing ratings as `NULL`

```sql
-- Jacob
INSERT INTO ratings (person_id, movie_id, rating)
SELECT p.person_id, m.movie_id, r.rating
FROM (VALUES
  ('Jacob','Dune: Part Two',5),
  ('Jacob','Barbie',4),
  ('Jacob','Oppenheimer',5),
  ('Jacob','Killers of the Flower Moon',NULL),
  ('Jacob','The Batman',4),
  ('Jacob','Spider-Man: Across the Spider-Verse',5)
) AS r(name,title,rating)
JOIN people p ON p.name = r.name
JOIN movies m ON m.title = r.title;

-- Alice
INSERT INTO ratings (person_id, movie_id, rating)
SELECT p.person_id, m.movie_id, r.rating
FROM (VALUES
  ('Alice','Dune: Part Two',4),
  ('Alice','Barbie',5),
  ('Alice','Oppenheimer',4),
  ('Alice','Killers of the Flower Moon',3),
  ('Alice','The Batman',NULL),
  ('Alice','Spider-Man: Across the Spider-Verse',5)
) AS r(name,title,rating)
JOIN people p ON p.name = r.name
JOIN movies m ON m.title = r.title;

-- Ben
INSERT INTO ratings (person_id, movie_id, rating)
SELECT p.person_id, m.movie_id, r.rating
FROM (VALUES
  ('Ben','Dune: Part Two',5),
  ('Ben','Barbie',3),
  ('Ben','Oppenheimer',5),
  ('Ben','Killers of the Flower Moon',4),
  ('Ben','The Batman',4),
  ('Ben','Spider-Man: Across the Spider-Verse',NULL)
) AS r(name,title,rating)
JOIN people p ON p.name = r.name
JOIN movies m ON m.title = r.title;

-- Chloe
INSERT INTO ratings (person_id, movie_id, rating)
SELECT p.person_id, m.movie_id, r.rating
FROM (VALUES
  ('Chloe','Dune: Part Two',NULL),
  ('Chloe','Barbie',5),
  ('Chloe','Oppenheimer',4),
  ('Chloe','Killers of the Flower Moon',NULL),
  ('Chloe','The Batman',5),
  ('Chloe','Spider-Man: Across the Spider-Verse',5)
) AS r(name,title,rating)
JOIN people p ON p.name = r.name
JOIN movies m ON m.title = r.title;

-- David
INSERT INTO ratings (person_id, movie_id, rating)
SELECT p.person_id, m.movie_id, r.rating
FROM (VALUES
  ('David','Dune: Part Two',4),
  ('David','Barbie',4),
  ('David','Oppenheimer',NULL),
  ('David','Killers of the Flower Moon',4),
  ('David','The Batman',3),
  ('David','Spider-Man: Across the Spider-Verse',4)
) AS r(name,title,rating)
JOIN people p ON p.name = r.name
JOIN movies m ON m.title = r.title;
```

## Quick SQL check

```sql
SELECT
  m.title,
  COUNT(r.rating) AS n_ratings,            -- counts non-NULL ratings
  ROUND(AVG(r.rating)::numeric, 2) AS avg_rating
FROM movies m
LEFT JOIN ratings r ON r.movie_id = m.movie_id
GROUP BY m.title
ORDER BY avg_rating DESC NULLS LAST;
```

Output:
```
                title                | n_ratings | avg_rating 
-------------------------------------+-----------+------------
 Spider-Man: Across the Spider-Verse |         4 |       4.75
 Dune: Part Two                      |         4 |       4.50
 Oppenheimer                         |         4 |       4.50
 Barbie                              |         5 |       4.20
 The Batman                          |         4 |       4.00
 Killers of the Flower Moon          |         3 |       3.67
(6 rows)
```

# Load data from PostgreSQL into a dataframe

## Install packages (run in RStudio Console)

```
install.packages(c("DBI", "RPostgres", "dplyr", "tidyr", "ggplot2"))
```

## Connect safely (password in environment variable)

In the terminal:

```bash
export PGPASSWORD='PostgreSQL_PASSWORD'
```

```{r}
library(DBI)
library(RPostgres)
library(dplyr)
library(tidyr)
library(ggplot2)

con <- dbConnect(
  RPostgres::Postgres(),
  dbname = "movie_ratings",
  host = "localhost",
  port = 5432,
  user = "postgres"
)
```
## Load ratings into R

```{r}
query <- "
SELECT
  p.name,
  m.title,
  m.release_year,
  r.rating
FROM ratings r
JOIN people p ON p.person_id = r.person_id
JOIN movies m ON m.movie_id = r.movie_id
ORDER BY m.title, p.name;
"

df <- dbGetQuery(con, query)
df |> head()
```

# Handle missing data + analysis

Missing-data strategy:  
- In SQL we stored “not seen / not rated” as `NULL`  
- In R, those become `NA`  
- For averages, we ignore `NA` using `na.rm = TRUE`  
- Report how many people rated each movie so averages aren’t misleading  

## Movie summary table

```{r}
movie_summary <- df |>
  group_by(title, release_year) |>
  summarise(
    n_possible = n(),                 # total people rows per movie (should be 5 here)
    n_rated = sum(!is.na(rating)),    # how many actually rated
    avg_rating = mean(rating, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(avg_rating))

movie_summary
```

## Simple visualization

```{r}
ggplot(movie_summary, aes(x = reorder(title, avg_rating), y = avg_rating)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Average Rating by Movie (ignoring missing ratings)",
    x = "Movie",
    y = "Average rating (1–5)"
  )
```

