---
title: "11-personalized-recommender"
format:
    html:
      toc: true
---

## Introduction

### Algorithm

**User-to-User Collaborative Filtering** (User-user CF) with Pearson correlation — recommend movies a user hasn't seen based on what similar-taste users rated highly.

1. Find your taste twins — look at all other users and measure how similarly they rate movies compared to you.
2. Borrow their opinions — for a movie you haven't seen, check what your taste twins thought of it.
3. Predict your rating — combine their ratings, giving more weight to users more similar to you.


### Approach

- **Input:** A 5-user x 6-movie rating matrix with 6 missing entries.
- **Build a rating matrix**: rows = users, columns = movies, cells = ratings (some missing).
- **Output:**
  - **Predict a missing rating**: take the target user's average rating, then nudge it up or down based on how similar users rated that movie (weighted by similarity).
  - **Recommend Top-N**: for each user, rank their *unseen* movies by predicted rating and return the top few.
- **Evaluate with Leave-One-Out CV**: hide one known rating at a time, predict it, compare to the real value, then report RMSE and MAE across all 24 ratings.

## Solution

### Load the data

```{r, message=FALSE}
library(knitr)
movie_ratings <- "https://raw.githubusercontent.com/qingquan-li/Data607/refs/heads/main/week3-assignment/data/movie_ratings.csv"
ratings <- read.csv(movie_ratings, check.names = FALSE, row.names = 1)
R <- as.matrix(ratings)
kable(R, caption = "Original rating matrix (NA = unrated)")
```

### User-user CF predictor

LLM prompt

> Write an R function predict_uu(R, u, i) that implements user-to-user collaborative filtering. R is a user×item matrix with NA for missing ratings. Use Pearson correlation on commonly-rated items as the similarity. Predict the rating as the target user's mean plus the similarity-weighted average of other users' mean-centered deviations on item i. Skip neighbors with fewer than 2 commonly-rated items, and fall back to the user's mean if no valid neighbors exist.

```{r}
predict_uu <- function(R, u, i) {
  u_mean <- mean(R[u, ], na.rm = TRUE)
  sims <- c(); devs <- c()
  for (v in setdiff(rownames(R), u)) {
    if (is.na(R[v, i])) next
    common <- !is.na(R[u, ]) & !is.na(R[v, ])
    if (sum(common) < 2) next
    a <- R[u, common] - mean(R[u, common])
    b <- R[v, common] - mean(R[v, common])
    denom <- sqrt(sum(a^2)) * sqrt(sum(b^2))
    if (denom == 0) next
    sims <- c(sims, sum(a * b) / denom)
    devs <- c(devs, R[v, i] - mean(R[v, ], na.rm = TRUE))
  }
  if (length(sims) == 0 || sum(abs(sims)) == 0) return(u_mean)
  u_mean + sum(sims * devs) / sum(abs(sims))
}
```

### Predicted ratings and Top-N recommendations

```{r}
P <- R
for (u in rownames(R)) for (i in colnames(R)) {
  if (is.na(R[u, i])) P[u, i] <- predict_uu(R, u, i)
}
kable(round(P, 2), caption = "Filled matrix (predictions replace NAs)")

topn <- function(u, n = 2) {
  mask <- is.na(R[u, ])
  if (!any(mask)) return("(no unseen items)")
  unseen <- colnames(R)[mask]
  preds <- setNames(P[u, unseen], unseen)   # keep names even if length 1
  preds <- sort(preds, decreasing = TRUE)
  k <- min(n, length(preds))
  paste0(names(preds)[1:k], " (", sprintf("%.2f", preds[1:k]), ")",
         collapse = "; ")
}
recs <- data.frame(User = rownames(R),
                   Top2 = sapply(rownames(R), topn, n = 2))
kable(recs, row.names = FALSE, caption = "Top-2 recommendations per user (drawn only from movies that user has not rated)")
```

### Evaluation: Leave-One-Out CV

```{r}
known <- which(!is.na(R), arr.ind = TRUE)
errs <- numeric(nrow(known))
for (k in seq_len(nrow(known))) {
  r <- known[k, 1]; c <- known[k, 2]
  R_tr <- R; R_tr[r, c] <- NA
  errs[k] <- R[r, c] - predict_uu(R_tr, rownames(R)[r], colnames(R)[c])
}
cat(sprintf("Held-out ratings: %d\nRMSE: %.3f\nMAE:  %.3f\n",
            length(errs), sqrt(mean(errs^2)), mean(abs(errs))))
```


## Conclusion

- **Held-out ratings: 24** — we ran the test 24 times, each time hiding one of the 24 known ratings (30 − 6 = 24 known ratings), predicting it from the remaining data, and comparing to the true value.
- **MAE: 0.667** — on average, the predicted rating is off by about **0.67 stars** (on a 1–5 scale). So if the true rating is 4, the model typically guesses somewhere between 3.3 and 4.7.
- **RMSE: 0.840** — similar to MAE, but squares the errors before averaging, so it **penalizes big misses more heavily**. RMSE > MAE means a few predictions were noticeably further off than the typical case.
- **Takeaway**: an average error under 1 star on a 1–5 scale is reasonable (acceptable) for such a tiny dataset (5 users, 6 movies) — the model is clearly learning preferences, not guessing randomly, but the small sample limits accuracy.