---
title:  "Project 2 — Data Transformations: GDP by Country"
author: "Qingquan Li"
date: 03/08/2026
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  echo:    true
  warning: false
---

## 1 — Data Source

This dataset contains **nominal GDP** (in millions USD) for 10 countries across six decades (1970–2020). The data originates from the Wikipedia page [List of countries by past and projected GDP (nominal)](https://en.wikipedia.org/wiki/List_of_countries_by_past_and_projected_GDP_(nominal)), which compiles figures from the **IMF World Economic Outlook** database.

The dataset is **untidy** because the year columns (1970, 1980, …) are values of a variable, not variable names.

## 2 — Data Structure Before Tidying

```{r load-libraries}
library(tidyr)
library(dplyr)
library(ggplot2)
```

```{r import-raw}
raw_gdp <- read.csv("https://raw.githubusercontent.com/qingquan-li/Data607/refs/heads/main/week6/data/dataset1_gdp.csv",
                     check.names = FALSE)
```

```{r preview-raw}
dim(raw_gdp)
head(raw_gdp)
str(raw_gdp)
```

The raw dataset has **10 rows** (one per country) and **7 columns** (Country + 6 year columns). Each year column stores GDP values — this is a classic wide format.

## 3 — Transformation Steps

```{r tidy-gdp}
tidy_gdp <- raw_gdp |>
  # Step 1: Reshape from wide to long format
  pivot_longer(
    cols      = -Country,
    names_to  = "year",
    values_to = "gdp_millions"
  ) |>
  # Step 2: Convert year from character to integer

  mutate(year = as.integer(year)) |>
  # Step 3: Rename columns to lowercase snake_case
  rename(country = Country) |>
  # Step 4: Remove any rows with missing GDP values
  drop_na(gdp_millions) |>
  # Step 5: Sort by country and year for readability
  arrange(country, year)
```

```{r preview-tidy}
head(tidy_gdp, 12)
str(tidy_gdp)
```

The tidy dataset now has three columns — `country`, `year`, and `gdp_millions` — with each row representing one observation (one country in one year).

## 4 — Analysis

**Goal:** Compare GDP growth trajectories across countries from 1970 to 2020.

### Summary Table

```{r summary-table}
gdp_summary <- tidy_gdp |>
  group_by(country) |>
  summarise(
    gdp_1970       = gdp_millions[year == 1970],
    gdp_2020       = gdp_millions[year == 2020],
    growth_factor  = round(gdp_2020 / gdp_1970, 1),
    .groups = "drop"
  ) |>
  arrange(desc(growth_factor))

gdp_summary
```

### Visualization

```{r gdp-plot, fig.width=10, fig.height=6}
ggplot(tidy_gdp, aes(x = year, y = gdp_millions / 1e6, color = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Nominal GDP by Country (1970–2020)",
    x     = "Year",
    y     = "GDP (Trillions USD)",
    color = "Country"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

## 5 — Conclusions

The **United States** maintained the largest GDP throughout the entire period, growing roughly 20-fold from \$1.07 trillion in 1970 to \$21.06 trillion in 2020.

However, the most dramatic growth story belongs to **China**, which grew over **159-fold** — from about \$93 billion to nearly \$14.7 trillion — overtaking Japan and Germany to become the world's second-largest economy.

**South Korea** also showed remarkable growth, which grew over 183-fold, rising from under \$9 billion to over \$1.6 trillion.

In contrast, **Brazil's** trajectory was uneven, peaking around 2010 before declining by 2020, illustrating how currency fluctuations and economic instability can affect nominal GDP rankings.
