---
title: "Project 2 — Data Transformations: Quarterly Revenue by Company"
author: "Qingquan Li"
date: 03/08/2026
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  echo: true
  warning: false
---

## 1 — Data Source

This dataset contains **quarterly revenue** (in billions USD) for five major technology companies (Apple, Microsoft, Amazon, Google, Meta) across eight quarters from Q1 2022 through Q4 2023. The figures were compiled from publicly available earnings reports.

Source: Company quarterly earnings reports (SEC filings / investor relations pages).

The dataset is **untidy** because the quarter columns (Q1_2022, Q2_2022, …) are values of a variable (quarter), not variable names.

## 2 — Data Structure Before Tidying

```{r load-libraries}
library(tidyr)
library(dplyr)
library(ggplot2)
```

```{r import-raw}
raw_rev <- read.csv("https://raw.githubusercontent.com/qingquan-li/Data607/refs/heads/main/week6/data/dataset3_revenue.csv",
                    check.names = FALSE)
```

```{r preview-raw}
dim(raw_rev)
head(raw_rev)
str(raw_rev)
```

The raw dataset has **5 rows** (one per company) and **9 columns** (Company + 8 quarter columns). Each quarter column stores a revenue figure — this is wide format.

## 3 — Transformation Steps

```{r tidy-rev}
tidy_rev <- raw_rev |>
  # Step 1: Reshape from wide to long format
  pivot_longer(
    cols      = -Company,
    names_to  = "quarter",
    values_to = "revenue_billions"
  ) |>
  # Step 2: Rename columns to lowercase snake_case
  rename(company = Company) |>
  # Step 3: Separate quarter into quarter label and year
  mutate(
    year       = as.integer(sub(".*_(\\d{4})$", "\\1", quarter)),
    qtr        = sub("_(\\d{4})$", "", quarter)
  ) |>
  # Step 4: Create an ordered factor for correct plotting order
  mutate(quarter = factor(quarter,
    levels = c("Q1_2022","Q2_2022","Q3_2022","Q4_2022",
               "Q1_2023","Q2_2023","Q3_2023","Q4_2023"),
    ordered = TRUE
  )) |>
  # Step 5: Remove any rows with missing values
  drop_na(revenue_billions) |>
  # Step 6: Sort by company and quarter
  arrange(company, quarter)
```

```{r preview-tidy}
head(tidy_rev, 10)
str(tidy_rev)
```

The tidy dataset now has columns `company`, `quarter`, `revenue_billions`, `year`, and `qtr`, with each row representing one observation (one company in one quarter).

## 4 — Analysis

**Goal:** Compare quarterly revenue trends across the five tech companies.

### Summary Table

```{r summary-table}
rev_summary <- tidy_rev |>
  group_by(company) |>
  summarise(
    avg_revenue = round(mean(revenue_billions), 1),
    min_revenue = min(revenue_billions),
    max_revenue = max(revenue_billions),
    yoy_growth  = round(
      (mean(revenue_billions[year == 2023]) /
       mean(revenue_billions[year == 2022]) - 1) * 100, 1
    ),
    .groups = "drop"
  ) |>
  arrange(desc(avg_revenue))

rev_summary
```

### Visualization

```{r rev-plot, fig.width=10, fig.height=6}
ggplot(tidy_rev, aes(x = quarter, y = revenue_billions, color = company, group = company)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Quarterly Revenue by Company (2022–2023)",
    x     = "Quarter",
    y     = "Revenue (Billions USD)",
    color = "Company"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )
```

## 5 — Conclusions

**Amazon** leads in total revenue, averaging about \$136 billion per quarter, driven by its diversified e-commerce and cloud business.

All five companies show a consistent **Q4 seasonal spike**, reflecting holiday-season consumer spending — this pattern is especially pronounced for Apple and Amazon.

Year-over-year, **Meta** showed the strongest growth at 15.6%.

**Microsoft** demonstrated steady, consistent growth across most quarters, likely fueled by increasing demand for its Azure cloud and AI services.

**Apple's** revenue remained relatively flat year-over-year, reflecting a maturing smartphone market.
