---
title:  "Project 2 — Data Transformations: Monthly Temperature by City"
author: "Qingquan Li"
date: 03/08/2026
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  echo:    true
  warning: false
---

## 1 — Data Source

This dataset contains **average monthly temperatures** (°F) for six major U.S. cities. The figures are based on NOAA 1991–2020 Climate Normals. The data was compiled into a wide-format CSV where each month is a separate column.

Source: [NOAA Climate Normals](https://www.ncei.noaa.gov/products/land-based-station/us-climate-normals)

The dataset is **untidy** because the month columns (Jan, Feb, …) are values of a variable (month), not variable names.

## 2 — Data Structure Before Tidying

```{r load-libraries}
library(tidyr)
library(dplyr)
library(ggplot2)
```

```{r import-raw}
raw_temp <- read.csv("https://raw.githubusercontent.com/qingquan-li/Data607/refs/heads/main/week6/data/dataset2_temperature.csv",
                     check.names = FALSE)
```

```{r preview-raw}
dim(raw_temp)
head(raw_temp)
str(raw_temp)
```

The raw dataset has **6 rows** (one per city) and **13 columns** (City + 12 month columns). Each month column stores a temperature value — classic wide format.

## 3 — Transformation Steps

```{r tidy-temp}
# Define month order for proper sorting
month_levels <- c("Jan","Feb","Mar","Apr","May","Jun",
                  "Jul","Aug","Sep","Oct","Nov","Dec")

tidy_temp <- raw_temp |>
  # Step 1: Reshape from wide to long format
  pivot_longer(
    cols      = -City,
    names_to  = "month",
    values_to = "avg_temp_f"
  ) |>
  # Step 2: Rename columns to lowercase snake_case
  rename(city = City) |>
  # Step 3: Convert month to an ordered factor for correct plotting order
  mutate(month = factor(month, levels = month_levels, ordered = TRUE)) |>
  # Step 4: Add a month number column for convenience
  mutate(month_num = as.integer(month)) |>
  # Step 5: Remove any rows with missing values
  drop_na(avg_temp_f) |>
  # Step 6: Sort by city and month
  arrange(city, month_num)
```

```{r preview-tidy}
head(tidy_temp, 12)
str(tidy_temp)
```

The tidy dataset now has columns `city`, `month`, `avg_temp_f`, and `month_num`, with each row representing one observation (one city in one month).

## 4 — Analysis

**Goal:** Compare seasonal temperature patterns across U.S. cities.

### Summary Table

```{r summary-table}
temp_summary <- tidy_temp |>
  group_by(city) |>
  summarise(
    min_temp  = min(avg_temp_f),
    max_temp  = max(avg_temp_f),
    mean_temp = round(mean(avg_temp_f), 1),
    range     = max_temp - min_temp,
    .groups   = "drop"
  ) |>
  arrange(desc(range))

temp_summary
```

### Visualization

```{r temp-plot, fig.width=10, fig.height=6}
ggplot(tidy_temp, aes(x = month, y = avg_temp_f, color = city, group = city)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Average Monthly Temperature by City",
    x     = "Month",
    y     = "Temperature (°F)",
    color = "City"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

## 5 — Conclusions

The analysis reveals distinct seasonal patterns across the six cities.

**Phoenix** has the highest summer peak at 95 °F and the widest temperature range (41 °F), reflecting its desert climate.

**Los Angeles** has the most stable temperatures year-round, with a range of only 18 °F, thanks to its coastal Mediterranean climate.

**Chicago** experiences the most extreme winters (26 °F in January) among the cities, while **Houston** stays relatively warm even in winter.

**Seattle** shows the smallest seasonal swing among the northern cities, with mild year-round temperatures due to its Pacific maritime influence.

These patterns are consistent with each city's geographic and climatic profile.
