---
title: "Week 9: Exploring the New York Times Article Search API"
format:
  html:
    toc: true
    theme: cosmo
    code-fold: false
editor: visual
---

## Introduction

**API Selected:** [New York Times Article Search API](https://developer.nytimes.com/docs/articlesearch-product/1/overview)

The Article Search API lets you query the full archive of NYT articles by keyword, date range, section, and other facets. Each result includes the headline, snippet, publication date, section, word count, and more.

**Research Question:** *How has New York Times coverage of "artificial intelligence" changed year over year from 2020 to 2025 --- particularly around the release of ChatGPT in late 2022 --- and which newspaper sections publish AI stories most frequently?*

**Planned Approach:**

1.  I will query the Article Search API for articles matching "artificial intelligence" across five individual years (2020--2025), collecting up to 20 articles per year (2 pages of 10 results).

2.  Then parse the JSON responses into a tidy data frame, visualize the year-over-year trend in article volume, and compare the top newspaper sections year by year to see how AI coverage shifted --- especially around the release of ChatGPT in late 2022.

Note: Due to API call limits, loading data may take 2-3 minutes.

```{r setup, message=FALSE, warning=FALSE}
library(httr2)
library(jsonlite)
library(tidyverse)
library(tidytext)
library(lubridate)
```

## Solution

### 1. Authentication

The API key is stored in the `.Renviron` file (one line: `NYT_API_KEY=your-key-here`) and retrieved at runtime so it is never hard-coded.

```{r auth}
api_key <- Sys.getenv("NYT_API_KEY")

if (nchar(api_key) == 0) {
  stop("NYT_API_KEY not found. Add it to your .Renviron file ",
       "(e.g., NYT_API_KEY=abc123) and restart R.")
}
```

### 2. Helper Function --- Querying the Article Search API

The Article Search API returns 10 articles per page and allows paging through results with the `page` parameter (0--99). We respect the API rate limit of 5 requests per minute by inserting a 12-second pause between calls.

```{r helper-function}
search_nyt <- function(query, begin_date, end_date, page = 0, api_key) {
  resp <- request("https://api.nytimes.com/svc/search/v2/articlesearch.json") |>
    req_url_query(
      q          = query,
      begin_date = begin_date,
      end_date   = end_date,
      page       = page,
      `api-key`  = api_key
    ) |>
    req_perform()

  resp |> resp_body_json()
}
```

### 3. Collecting Data (2020--2025)

We loop over each year and pull 2 pages (up to 20 articles) per year.

Note: Due to API call limits, loading data may take 2-3 minutes.

```{r collect-data, cache=TRUE}
years <- 2020:2025
all_articles <- list()

for (yr in years) {
  begin <- paste0(yr, "0101")
  end   <- paste0(yr, "1231")

  for (pg in 0:1) {
    Sys.sleep(12)  # free tier allows 5 req/min -> 12s between calls
    message("Fetching year ", yr, " page ", pg + 1, " of 2...")

    result <- NULL
    for (attempt in 1:3) {
      result <- tryCatch(
        search_nyt("artificial intelligence", begin, end, pg, api_key),
        error = function(e) {
          if (grepl("429", e$message)) {
            message("  Rate limited -- waiting 60s then retrying (attempt ", attempt, "/3)...")
            Sys.sleep(60)
            return(NULL)
          }
          message("Error on year ", yr, " page ", pg, ": ", e$message)
          return(NULL)
        }
      )
      if (!is.null(result)) break
    }

    if (is.null(result)) next

    docs <- result$response$docs
    if (length(docs) == 0) break  # no more results for this year

    all_articles <- c(all_articles, docs)
  }
}

message("Total raw articles collected: ", length(all_articles))
```

### 4. Parsing JSON into a Tidy Data Frame

Each element of `all_articles` is a nested list. We extract the fields we need and handle missing values explicitly.

```{r parse-json}
articles_df <- map_dfr(all_articles, function(doc) {
  tibble(
    headline     = doc$headline$main       %||% NA_character_,
    snippet      = doc$snippet             %||% NA_character_,
    section      = doc$section_name        %||% NA_character_,
    pub_date     = doc$pub_date            %||% NA_character_,
    word_count   = doc$word_count          %||% NA_integer_,
    web_url      = doc$web_url             %||% NA_character_,
    news_desk    = doc$news_desk           %||% NA_character_,
    type_of_material = doc$type_of_material %||% NA_character_,
    document_type = doc$document_type      %||% NA_character_
  )
})

glimpse(articles_df)
```

### 5. Data Cleaning

```{r clean}
articles_clean <- articles_df |>
  mutate(
    pub_date   = as_date(pub_date),
    year       = year(pub_date),
    word_count = as.integer(word_count),
    # Standardise blank / NA sections to "Unknown"
    section    = if_else(is.na(section) | section == "", "Unknown", section)
  ) |>
  # Drop any accidental duplicates (same URL = same article)
  distinct(web_url, .keep_all = TRUE)

cat("Clean rows:", nrow(articles_clean), "\n")
cat("Date range:", as.character(min(articles_clean$pub_date, na.rm = TRUE)),
    "to", as.character(max(articles_clean$pub_date, na.rm = TRUE)), "\n")
```

**Data-cleaning notes:**

-   **Nested fields:** The headline lives inside a nested `headline$main` subfield; we extracted it directly during parsing.
-   **Missing values:** `section_name`, `word_count`, and `news_desk` are occasionally `NULL` in the API response. We coalesce these to `NA` using the `%||%` operator and later recode blank sections as "Unknown."
-   **Duplicates:** The Article Search API can return the same article on different query pages. We de-duplicate on `web_url`.
-   **Date conversion:** `pub_date` arrives as an ISO-8601 string; we convert it to a proper `Date` column and derive a `year` column for grouping.

### 6. Analysis & Visualization

#### Preview of the Final Data Frame

```{r preview}
articles_clean |>
  select(headline, section, pub_date, word_count) |>
  slice_head(n = 10) |>
  knitr::kable()
```

#### Top Sections for AI Coverage --- Year by Year

Facet by year so we can directly compare which sections dominated AI coverage before and after ChatGPT (released Nov 2022).

```{r plot-sections-by-year, fig.width=10, fig.height=8}
# Count articles per section per year, keep top 5 sections within each year
sections_by_year <- articles_clean |>
  count(year, section, name = "n") |>
  group_by(year) |>
  slice_max(n, n = 5, with_ties = FALSE) |>
  ungroup()

ggplot(sections_by_year,
       aes(x = reorder_within(section, n, year), y = n, fill = factor(year))) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.2, size = 3) +
  coord_flip() +
  facet_wrap(~ year, scales = "free_y", ncol = 1) +
  scale_x_reordered() +
  labs(
    title = "Top 5 NYT Sections for AI Articles, by Year (2020-2025)",
    subtitle = "Comparing section mix before and after ChatGPT (Nov 2022)",
    x = NULL, y = "Number of Articles"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold", size = 11))
```

## Conclusion

This analysis queried the NYT Article Search API for articles about artificial intelligence published from 2020 to 2025, allowing us to compare coverage before and after the release of ChatGPT in late November 2022. A few findings stand out:

1.  **The ChatGPT inflection point.** We expect to see relatively flat coverage in 2020--2022, followed by a sharp jump in 2023 and 2025. ChatGPT brought AI from a niche technology beat into mainstream public consciousness almost overnight.

2.  **Section concentration.** By comparing the top sections year by year, we can see whether AI coverage was once confined to Technology and Science and then spread to sections like U.S. and Business after ChatGPT brought AI into mainstream public discourse.
