---
title: "10a-sentiment-analysis-with-text-mining-with-r"
format:
    html:
      toc: true
      embed-resources: true
    pdf:
      pdf-engine: pdflatex
      geometry: margin=1in
---

# Introduction

This assignment reproduces the sentiment analysis example from Chapter 2 of
*Text Mining with R* (Silge & Robinson, 2017) and extends it with a new
corpus and an additional sentiment lexicon.

**Citation:** Silge, J., & Robinson, D. (2017). *Text Mining with R: A Tidy
Approach*. O'Reilly Media. Chapter 2: Sentiment Analysis with Tidy Data.
Available at https://www.tidytextmining.com/sentiment. The code in Part 1
is adapted directly from that chapter.

# Setup

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidytext)
library(janeaustenr)
library(gutenbergr)
```

# Part 1: Reproducing the Base Example

The Chapter 2 example tokenizes Jane Austen's novels, joins each token with
the `bing` lexicon, and plots net sentiment (positive minus negative) across
80-line sections of each book.

```{r reproduce, message=FALSE, warning=FALSE}
tidy_books <- austen_books() %>%
  group_by(book) %>%
  mutate(
    linenumber = row_number(),
    chapter = cumsum(str_detect(text, 
                                regex("^chapter [\\divxlc]", 
                                      ignore_case = TRUE)))) %>%
  ungroup() %>%
  unnest_tokens(word, text)

jane_austen_sentiment <- tidy_books %>%
  inner_join(get_sentiments("bing")) %>%
  count(book, index = linenumber %/% 80, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>% 
  mutate(sentiment = positive - negative)

ggplot(jane_austen_sentiment, aes(index, sentiment, fill = book)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~book, ncol = 2, scales = "free_x")
```

This reproduces Figure 2.2 of the book.

# Part 2: Extension

## New Corpus

We pull *The Adventures of Sherlock Holmes* (Arthur Conan Doyle, 1892) from
Project Gutenberg (ID 1661).

```{r corpus, message=FALSE, warning=FALSE}
holmes <- gutenberg_download(1661, mirror = "https://gutenberg.pglaf.org/")

tidy_holmes <- holmes |>
  mutate(linenumber = row_number()) |>
  unnest_tokens(word, text)
```

## Additional Lexicon: Loughran-McDonald

The Loughran-McDonald lexicon (Loughran & McDonald, 2011) was developed for
financial and regulatory text (10-K filings, earnings releases). It is
distributed with `tidytext` but is *not* used in the Chapter 2 example, so
we treat it as our additional lexicon. It classifies words into six
categories: `positive`, `negative`, `litigious`, `uncertainty`,
`constraining`, and `superfluous`. For comparability we keep only the
positive/negative categories.

**Citation:** Loughran, T., & McDonald, B. (2011). When is a liability not
a liability? Textual analysis, dictionaries, and 10-Ks. *Journal of
Finance*, 66(1), 35-65.

## Comparing Four Lexicons on Sherlock Holmes

We score net sentiment over 80-line sections using `bing`, `nrc`, `afinn`,
and `loughran`. For `afinn` we binarize the integer scores (non-negative =
positive) so all four are on the same footing.

```{r compare, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
binary <- function(lex) {
  s <- get_sentiments(lex)
  if (lex == "afinn") {
    s |> transmute(word, sentiment = if_else(value >= 0, "positive", "negative"))
  } else {
    s |> filter(sentiment %in% c("positive", "negative"))
  }
}

score <- function(lex) {
  tidy_holmes |>
    inner_join(binary(lex), by = "word", relationship = "many-to-many") |>
    count(index = linenumber %/% 80, sentiment) |>
    pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) |>
    mutate(sentiment = positive - negative, lexicon = lex)
}

holmes_compare <- bind_rows(
  score("bing"),
  score("nrc"),
  score("afinn"),
  score("loughran")
)

ggplot(holmes_compare, aes(index, sentiment, fill = lexicon)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~lexicon, ncol = 2, scales = "free_y") +
  labs(title = "Sentiment through 'The Adventures of Sherlock Holmes'",
       x = "Narrative time (80-line sections)",
       y = "Net sentiment")
```

## Summary Statistics

```{r summary, message=FALSE}
holmes_compare |>
  group_by(lexicon) |>
  summarise(
    mean_net = round(mean(sentiment), 2),
    sd_net   = round(sd(sentiment), 2),
    pct_neg  = round(mean(sentiment < 0) * 100, 1)
  )
```

## How Do the Results Differ?

- **nrc** shows a strong positive bias because its positive category
  contains many common everyday words. Holmes looks overwhelmingly
  positive under `nrc`, even in chapters built around murder and
  blackmail.
- **bing** is more balanced and picks up the darker plot points that
  `nrc` washes out. This matches what Silge and Robinson observe in
  Section 2.3.
- **afinn**, once binarized, looks similar to `bing` in shape. The raw
  integer weights (dropped here for comparability) would make it more
  sensitive to a single strongly negative word per section.
- **loughran** was trained on corporate filings, so most of its
  vocabulary (`litigation`, `restructuring`, `impairment`) does not
  appear in a Victorian detective story. Its signal is sparse and mostly
  negative, driven by crime-adjacent overlap words like `crime`,
  `guilty`, and `suspicious`. This illustrates a key point: sentiment
  lexicons are domain-specific, and applying a finance lexicon to
  fiction under-counts sentiment and misses most of the emotional
  texture of the text.

The broader takeaway is the same one the book makes for Jane Austen: the
choice of lexicon materially changes the story you tell about a text, so
comparing multiple lexicons is more informative than trusting any single
one.


