---
title: "10b-more-json-practice-with-nobelprize-org"
format:
  html:
    toc: true
  pdf:
    latex-engine: pdflatex
execute:
  warning: false
  message: false
---

## Introduction

This assignment uses the public Nobel Prize API (v1) at
`http://api.nobelprize.org/v1/`. Two JSON endpoints are consumed:

- `prize.json` — one record per prize (year + category) with nested laureates
- `laureate.json` — one record per laureate with nested prizes and affiliations

Both are pulled, flattened into tidy data frames, and used to answer four
questions below.

```{r setup}
library(tidyverse)
library(jsonlite)

# Pull both endpoints once
prizes_raw    <- fromJSON("http://api.nobelprize.org/v1/prize.json",
                          simplifyVector = FALSE)
laureates_raw <- fromJSON("http://api.nobelprize.org/v1/laureate.json",
                          simplifyVector = FALSE)
```

### Tidy the prize data

Each prize has a list of laureates, so we unnest it into one row per
laureate-prize.

```{r tidy-prizes}
prizes_df <- tibble(prize = prizes_raw$prizes) |>
  unnest_wider(prize) |>
  mutate(year = as.integer(year)) |>
  unnest_longer(laureates) |>       # one row per laureate on each prize
  unnest_wider(laureates) |>
  filter(!is.na(id))                # drop years with no laureates (e.g., war years)

glimpse(prizes_df)
```

### Tidy the laureate data

```{r tidy-laureates}
laureates_df <- tibble(l = laureates_raw$laureates) |>
  unnest_wider(l) |>
  mutate(id = as.character(id))

# One row per (laureate, prize) with affiliation info for Q4
lp_df <- laureates_df |>
  select(id, firstname, surname, gender, bornCountry, prizes) |>
  unnest_longer(prizes) |>
  unnest_wider(prizes, names_sep = "_") |>
  mutate(prize_year = as.integer(prizes_year))

glimpse(lp_df)
```

---

## Question 1: Which category has awarded the most Nobel Prizes?

A simple count of laureates by category.

```{r q1}
q1 <- prizes_df |>
  count(category, sort = TRUE)
q1
```

```{r q1-plot, fig.height=3}
ggplot(q1, aes(reorder(category, n), n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(x = NULL, y = "Laureates", title = "Nobel laureates by category")
```

---

## Question 2: How has the number of laureates per decade changed over time?

```{r q2, fig.height=3}
q2 <- prizes_df |>
  mutate(decade = (year %/% 10) * 10) |>
  count(decade)

ggplot(q2, aes(decade, n)) +
  geom_line() + geom_point() +
  labs(x = "Decade", y = "Laureates",
       title = "Nobel laureates per decade")
q2
```

The count rises over time, reflecting the modern practice of splitting
prizes among multiple laureates.

---

## Question 3: What is the gender distribution within each category?

This joins laureate-level gender with prize-level category.

```{r q3}
# Join gender from laureates_df onto prizes_df by laureate id
q3 <- prizes_df |>
  mutate(id = as.character(id)) |>
  left_join(select(laureates_df, id, gender), by = "id") |>
  filter(gender %in% c("male", "female")) |>     # drop organizations
  count(category, gender) |>
  group_by(category) |>
  mutate(pct_female = round(100 * n[gender == "female"] / sum(n), 1)) |>
  ungroup()
q3
```

```{r q3-plot, fig.height=3}
ggplot(q3, aes(category, n, fill = gender)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Share", title = "Gender share by category") +
  coord_flip()
```

Peace and Literature have the highest share of female laureates; Physics
and Economics have the lowest.

---

## Question 4: Which country has "lost" the most laureates?

The advanced question: find laureates **born** in one country but
**affiliated** with a different country at the time of the award. This
requires unnesting prize affiliations and comparing against born country.

```{r q4}
# Unnest affiliations inside each prize
brain_drain <- lp_df |>
  select(id, bornCountry, prize_year, prizes_category, prizes_affiliations) |>
  unnest_longer(prizes_affiliations) |>
  unnest_wider(prizes_affiliations, names_sep = "_") |>
  rename(affil_country = prizes_affiliations_country) |>
  filter(!is.na(bornCountry),
         !is.na(affil_country),
         nzchar(bornCountry),
         nzchar(affil_country),
         bornCountry != affil_country) |>
  distinct(id, bornCountry, affil_country)       # one row per laureate move

q4 <- brain_drain |>
  count(bornCountry, sort = TRUE) |>
  slice_head(n = 10)
q4
```

```{r q4-plot, fig.height=3}
ggplot(q4, aes(reorder(bornCountry, n), n)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  labs(x = NULL, y = "Laureates lost",
       title = "Top 10 countries losing Nobel laureates to other countries")
```

Germany (Germany and Germany (now Poland)) tops the list — largely because many scientists born in the
pre-war German Empire were affiliated with U.S. or U.K. institutions by
the time they won, reflecting the 20th-century migration of scientific
talent.

---

## Summary

| # | Question | Answer |
|---|----------|--------|
| 1 | Most-awarded category | Medicine / Physics (largest counts) |
| 2 | Laureates per decade | Rising trend, peaking in recent decades |
| 3 | Gender balance | Peace & Literature most balanced; Physics least |
| 4 | Biggest "brain drain" | Germany |
