---
title: "project1 - Chess Tournament Parsing"
format: html
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

In this project, I will transform the semi-structured chess tournament text file into a clean CSV table containing:

- Player Name  
- Player State  
- Total Points  
- Player Pre-Rating  
- Average Pre-Rating of Opponents  

**Planned approach**

1. Read the raw text file line-by-line.  
2. Remove divider/header lines and group the remaining lines into 3-line "player blocks" (each player appears as 3 lines in the file).  
3. Parse each block to extract: pair number, name, points, state, and pre-rating.
4. From the round results line, extract each opponent’s *pair number* (ignoring byes/half-point byes/withdrawals where no opponent number exists).
5. Join opponent pair numbers back to the players table to get opponents' pre-ratings, then compute the average.


**Anticipated data challenges**

- The file is **fixed-width-ish** rather than a clean delimited format, so parsing relies on patterns.  
- Some rounds contain entries like **H, U, B, X** (half-point bye, unplayed, bye, forfeit) where an opponent number may be missing.  
- Ratings can include suffixes like **P** (provisional), e.g., `1641P17`, so I need to extract only the numeric rating reliably.  


# Solution

```{r}
# Load packages
# dplyr   -> data wrangling (select, mutate, join, summarize)
# stringr -> string/text parsing
# purrr   -> list operations (map, keep, discard)
# tidyr   -> unnest list-columns
library(dplyr)
library(stringr)
library(purrr)
library(tidyr)
library(tibble)
```

## 1. Read the raw text file

```{r}
path <- "https://raw.githubusercontent.com/qingquan-li/Data607/refs/heads/main/week4/tournamentinfo.txt"

# Read the file line by line (each row becomes one string)
raw_lines <- readLines(path, warn = FALSE)

# Preview the first few raw lines
head(raw_lines, 10)
```

## 2. Clean the lines

AI Prompt: Remove header rows, dashed divider rows, and blank rows from the attached tournamentinfo.txt file using R so only the player-related rows remain. Please add any necessary comments.

```{r}
lines <- raw_lines |>
  # Remove carriage return characters (sometimes appear in Windows text files)
  str_replace_all("\r", "") |>
  
  # Remove divider lines made of dashes ----------------------------
  discard(~ str_detect(.x, "^-{5,}$")) |>
  
  # Remove the main header row that starts with "Pair | Player Name ..."
  discard(~ str_detect(.x, "^\\s*Pair\\s*\\|")) |>
  
  # Remove the second header row that starts with "Num | USCF ID ..."
  discard(~ str_detect(.x, "^\\s*Num\\s*\\|")) |>
  
  # Remove blank/empty lines
  discard(~ str_detect(.x, "^\\s*$"))

# Check how many lines remain after cleaning
length(lines)

# Optional: print cleaned lines with row numbers if you want to inspect them
tibble(i = seq_along(lines), line = lines)
```

## 3. Build player blocks safely (3 lines per player)


```{r}
# IMPORTANT:
# Instead of assuming all remaining lines are perfectly grouped by 3,
# we find the start of each player block using a pattern.
#
# A player block starts with a line like:
# "  1 | GARY HUA ... "
# " 12 | SOME PLAYER ... "
#
# So we search for lines that begin with:
# optional spaces + digits + "|" 
start_idx <- which(str_detect(lines, "^\\s*\\d+\\s*\\|"))

# Create a list of blocks.
# Each block should contain:
#   line 1 = pair number, player name, points, round results
#   line 2 = state, USCF ID, pre/post rating
#   line 3 = extra info (we usually don't need much from this line)
blocks <- map(start_idx, ~ lines[.x:(.x + 2)])

# Keep only complete blocks (safety check)
# This avoids errors if the last block is incomplete for any reason
blocks <- keep(blocks, ~ length(.x) == 3)

# How many players/blocks did we detect?
length(blocks)

# Optional: inspect the first player block
blocks[[1]]
```


## 4. Parse one player block into structured columns

```{r}
# This function takes ONE 3-line block and extracts:
# - pair number
# - player name
# - state
# - total points
# - pre-rating
# - opponent pair numbers (for all rounds that have an opponent number)
parse_block <- function(block3) {
  
  # First line contains pair number, name, points, round results
  l1 <- block3[[1]]
  
  # Second line contains state and rating info
  l2 <- block3[[2]]
  
  # --- Extract pair number ---
  # Example: "  1 | GARY HUA ... " -> pair_num = 1
  pair_num <- str_extract(l1, "^\\s*\\d+") |> 
    str_squish() |> 
    as.integer()
  
  # --- Extract player name ---
  # Split line by "|" and take the 2nd field
  # Example fields:
  # [1] pair number, [2] player name, [3] points, [4+] rounds
  fields_l1 <- str_split(l1, "\\|", simplify = TRUE)
  
  name <- fields_l1[1, 2] |>
    str_squish() |>      # remove extra spaces
    str_to_title()       # convert "GARY HUA" -> "Gary Hua"
  
  # --- Extract total points ---
  # Usually the 3rd pipe-separated field
  total_points <- fields_l1[1, 3] |>
    str_squish() |>
    as.numeric()
  
  # --- Extract state ---
  # State is typically the first 2-letter code on line 2 (e.g., ON, MI, NY)
  player_state <- str_extract(l2, "^\\s*[A-Z]{2}") |>
    str_squish()
  
  # --- Extract pre-rating ---
  # Line 2 has something like "R: 1794   ->1817"
  # We want the FIRST rating (pre-tournament rating), i.e., 1794
  pre_rating <- str_extract(l2, "R:\\s*\\d+") |>
    str_extract("\\d+") |>
    as.integer()
  
  # --- Extract opponent pair numbers from round results ---
  # Round results in line 1 look like:
  # W  39 | L   4 | D  12 ...
  #
  # We extract only results that include W/L/D + an opponent number.
  # This automatically ignores H, U, B, X (bye/unplayed/etc.) because
  # they usually do not include a valid opponent number.
  opp_nums <- str_extract_all(l1, "\\b[WLD]\\s*\\d+\\b")[[1]] |>
    str_extract("\\d+") |>
    as.integer()
  
  # Return one row (as a tibble)
  tibble(
    pair_num = pair_num,
    player_name = name,
    player_state = player_state,
    total_points = total_points,
    pre_rating = pre_rating,
    opp_pair_nums = list(opp_nums)  # list-column because each player has multiple opponents
  )
}
```


## 5. Parse all players

```{r}
# Apply parse_block() to every 3-line block and combine into one data frame
players <- map_dfr(blocks, parse_block)

# Preview parsed player table
players
```


## 6. Compute average pre-rating of opponents

```{r}
# We need the average PRE-rating of each player's opponents.
#
# Strategy:
# 1) Create a lookup table: pair_num -> pre_rating
# 2) Expand (unnest) the opponent lists so each opponent is one row
# 3) Join to get each opponent's pre_rating
# 4) Group by player and compute the average

# Step 1: lookup table (player number -> pre-rating)
rating_lookup <- players |>
  select(pair_num, pre_rating)

# Step 2-4: expand opponents, join ratings, compute average
opponent_avgs <- players |>
  select(pair_num, opp_pair_nums) |>
  
  # Turn list-column into regular rows:
  # one player with 7 opponents becomes 7 rows
  unnest(opp_pair_nums, keep_empty = TRUE) |>
  
  # Rename for clarity: this is the opponent's pair number
  rename(opp_pair_num = opp_pair_nums) |>
  
  # Join to get opponent's pre-rating
  # left_join keeps all rows even if some opponent number is missing
  left_join(rating_lookup, by = c("opp_pair_num" = "pair_num")) |>
  
  # Now compute average opponent pre-rating for each player
  group_by(pair_num) |>
  summarize(
    avg_opp_pre_rating = mean(pre_rating, na.rm = TRUE),  # ignore NA values
    games_counted = sum(!is.na(pre_rating)),              # how many opponent ratings were used
    .groups = "drop"
  )

# Preview the computed averages
opponent_avgs
```

## 7. Build the final required output table

```{r}
# Join the opponent averages back to the player info and format the final result
final_df <- players |>
  select(player_name, player_state, total_points, pre_rating, pair_num) |>
  left_join(opponent_avgs, by = "pair_num") |>
  mutate(
    # Round to nearest integer (the project example uses whole number)
    avg_opp_pre_rating = round(avg_opp_pre_rating)
  ) |>
  select(
    `Player Name` = player_name,
    `Player State` = player_state,
    `Total Number of Points` = total_points,
    `Player's Pre-Rating` = pre_rating,
    `Average Pre Chess Rating of Opponents` = avg_opp_pre_rating
  )

# Show final output
final_df
```

## 8. Validate the first player (Gary Hua)

```{r}
# The project prompt says Gary Hua should be:
# Gary Hua, ON, 6.0, 1794, 1605
final_df |> filter(`Player Name` == "Gary Hua")
```


## 9. Export the CSV file

```{r}
# Write the result to CSV
output_file <- "./chess_tournament_players.csv"
write.csv(final_df, output_file, row.names = FALSE)

# Print the path so you know where the file is
output_file
```


# Conclusion

This project demonstrates how to turn a semi-structured text file into a database-friendly CSV using string parsing and tidy data workflows.

Key insights / findings:

- Even though the tournament file looks like a table, it behaves more like a fixed-format report; grouping lines into consistent player blocks (3 lines per player) makes parsing reliable.

- The main complication is handling rounds that don’t list an opponent number (byes/withdrawals/forfeits). The cleanest method is to compute averages only from rounds where an opponent number exists and can be matched to a real player.

- Once opponent pair numbers are extracted, the rest is standard data wrangling: join to get opponent ratings and summarize with a mean.

The final CSV can be imported directly into a SQL table for further analysis (e.g., comparing performance vs. opponent strength).

