---
title: "3a-global-baseline-estimate"
format: html
---

# Introduction

I use the survey data ([movie-ratings](https://raw.githubusercontent.com/qingquan-li/Data607/refs/heads/main/week3-assignment/data/movie_ratings.csv)) to build a simple recommender using the Global Baseline Estimate method. I'll compute three key components:\
1. the global mean rating across all observed ratings,\
2. each user's bias (how a user’s average rating differs from the global mean),\
3. each movie's bias (how a movie's average rating differs from the global mean).\
Using the spreadsheet's algorithm, I will predict an unseen rating with `predicted(u, m) = global_mean + user_bias(u) + movie_bias(m)`.

Finally, for each user, I will generate predictions for the movies they haven't rated and recommend the movie(s) with the highest predicted score.

# Body

## Load the ratings data (CSV)

The file `movie_ratings.csv` is a wide “ratings matrix” where rows are people and columns are movies. Missing values mean the person has not seen (or did not rate) the movie.

```{r}
library(tidyverse)

data_url <- "https://raw.githubusercontent.com/qingquan-li/Data607/refs/heads/main/week3-assignment/data/movie_ratings.csv"

ratings_wide <- read_csv(data_url, show_col_types = FALSE)

glimpse(ratings_wide)
ratings_wide
```

## Reshape to long format

For analysis, it is easier to work in a long format with one row per (user, movie, rating).

```{r}
ratings_long <- ratings_wide %>%
  rename(user = Person) %>%
  pivot_longer(
    cols = -user,
    names_to = "movie",
    values_to = "rating"
  )

ratings_long %>% sample_n(10)
```

### Missing data strategy

-   Missing ratings are represented as `NA`.
-   When computing averages, we exclude `NA` values so they do not distort the mean.
-   Predictions are generated only for movies a user has not rated (`rating` is `NA`).

```{r}
ratings_long %>% summarise(
  total_rows = n(),
  missing_ratings = sum(is.na(rating)),
  observed_ratings = sum(!is.na(rating))
)
```

## Global Baseline Estimate components

### 1) Global mean rating

This is the average of all observed ratings across all users and movies.

```{r}
global_mean <- ratings_long %>%
  summarise(mu = mean(rating, na.rm = TRUE)) %>%
  pull(mu)

global_mean
```

### 2) User bias

A user bias measures how a user rates compared to the global mean:

```{r}
user_bias <- ratings_long %>%
  group_by(user) %>%
  summarise(user_avg = mean(rating, na.rm = TRUE), n_ratings = sum(!is.na(rating)), .groups = "drop") %>%
  mutate(b_user = user_avg - global_mean) %>%
  arrange(desc(b_user))

user_bias
```

### 3) Movie bias

A movie bias measures how a movie is rated compared to the global mean:

```{r}
movie_bias <- ratings_long %>%
  group_by(movie) %>%
  summarise(movie_avg = mean(rating, na.rm = TRUE), n_ratings = sum(!is.na(rating)), .groups = "drop") %>%
  mutate(b_movie = movie_avg - global_mean) %>%
  arrange(desc(b_movie))

movie_bias
```

## Predict missing ratings and recommend movies

The Global Baseline Estimate prediction is:

```{r}
predictions <- ratings_long %>%
  left_join(user_bias %>% select(user, b_user), by = "user") %>%
  left_join(movie_bias %>% select(movie, b_movie), by = "movie") %>%
  mutate(pred_rating = global_mean + b_user + b_movie)

predictions
```

### Recommendations (top unseen movie per user)

```{r}
recommendations <- predictions %>%
  filter(is.na(rating)) %>%                 # only movies the user has NOT rated
  group_by(user) %>%
  slice_max(pred_rating, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(user, movie, pred_rating) %>%
  arrange(desc(pred_rating))

recommendations
```

# Conclusions

Using the Global Baseline Estimate method, I generated movie recommendations by combining the global average rating with a user bias and a movie bias. This provides a simple baseline recommender that works well with survey data, and it handles missing ratings naturally by predicting only for unseen movies.
