---
title: "3b-window-functions"
format: html
---

# Approach

To practice Window Functions (in dplyr), I will use a cryptocurrency time-series dataset with daily end-of-day prices for at least two coins (e.g., BTC-USD and ETH-USD) starting from January 1, 2022. I will store the data in CSV files with columns such as date, symbol, and close_price, ensuring one row per (date, symbol). Then I will write queries using window functions partitioned by symbol and ordered by date to compute:\
1. the year-to-date (YTD) average close price for each date (resetting at the start of each calendar year),\
2. a 6-day moving average using a rolling window (the last 6 dates for each coin).\
I will summarize the output in a small table or plot.

# Download BTC + ETH data

I use `tidyquant::tq_get()` to fetch daily prices from Yahoo Finance for multiple symbols.

```{r}
library(tidyquant)
library(dplyr)
library(readr)
library(lubridate)
library(slider)
library(ggplot2)

symbols <- c("BTC-USD", "ETH-USD")

crypto <- tq_get(
  symbols,
  get  = "stock.prices",
  from = "2022-01-01",
  to   = Sys.Date()
)

# Save the raw download
dir.create("data", showWarnings = FALSE)
write_csv(crypto, "data/crypto_btc_eth_daily.csv")

crypto %>% glimpse()
crypto %>% count(symbol)
```

# Window functions in dplyr: YTD average + 6-day moving average

## Year-to-date average (YTD)

To compute a YTD average that restarts each year:

-   create a `year` column from the date

-   group by `symbol` and `year` - apply `cummean(close)` (cumulative mean) within each group

## 6-day moving average

To compute a rolling 6-day moving average for each crypto:

-   group by `symbol`

-   order by date

-   take the mean of the current day and previous 5 days using `slider::slide_dbl(...)`

```{r}
crypto_windows <- crypto %>%
  arrange(symbol, date) %>%
  mutate(year = year(date)) %>%
  group_by(symbol, year) %>%
  mutate(
    ytd_avg_close = cummean(close)
  ) %>%
  group_by(symbol) %>%
  mutate(
    ma_6_close = slide_dbl(close, mean, .before = 5, .complete = TRUE)
  ) %>%
  ungroup()

crypto_windows %>%
  select(symbol, date, close, ytd_avg_close, ma_6_close) %>%
  tail(12)
```

## Save the enriched dataset

```{r}
write_csv(crypto_windows, "data/crypto_btc_eth_with_windows.csv")
file.exists("data/crypto_btc_eth_with_windows.csv")
```

# Quick visual check

A moving average should be smoother than the raw close price.

```{r}
crypto_windows %>%
  filter(symbol == "BTC-USD") %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = close, color = "Close"), alpha = 0.6) +
  geom_line(aes(y = ma_6_close, color = "6-Day MA")) +
  scale_color_manual(values = c("Close" = "steelblue", "6-Day MA" = "orange")) +
  labs(
    title = "BTC-USD: Close vs 6-Day Moving Average",
    x = "Date",
    y = "Price (USD)",
    color = "Series"
  )

```

```{r}
crypto_windows %>%
  filter(symbol == "ETH-USD") %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = close, color = "Close"), alpha = 0.6) +
  geom_line(aes(y = ma_6_close, color = "6-Day MA")) +
  scale_color_manual(values = c("Close" = "steelblue", "6-Day MA" = "orange")) +
  labs(
    title = "ETH-USD: Close vs 6-Day Moving Average",
    x = "Date",
    y = "Price (USD)",
    color = "Series"
  )
```

# Conclusions

Using daily cryptocurrency prices for BTC-USD and ETH-USD since January 1, 2022, I applied window-style operations in dplyr to compute **year-to-date average close prices** and **6-day moving averages** for each symbol. The YTD average provides a running baseline that resets each year, while the 6-day moving average smooths short-term noise and makes trends easier to interpret.

To extend this work, I could compare multiple window sizes (7/14/30 days), compute the same features on daily returns rather than raw prices, and add data-quality checks (duplicates, missing dates, extreme outliers) to ensure rolling calculations are based on a clean and consistent time series.


---


# Optional: 30-day moving average

```{r}
crypto_windows_30 <- crypto %>%
  arrange(symbol, date) %>%
  mutate(year = year(date)) %>%
  group_by(symbol, year) %>%
  mutate(
    ytd_avg_close = cummean(close)
  ) %>%
  group_by(symbol) %>%
  mutate(
    ma_30_close = slide_dbl(close, mean, .before = 29, .complete = TRUE)
  ) %>%
  ungroup()
```


Quick visual check:  
A moving average should be smoother than the raw close price.

```{r}
crypto_windows_30 %>%
  filter(symbol == "BTC-USD") %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = close, color = "Close"), alpha = 0.6) +
  geom_line(aes(y = ma_30_close, color = "30-Day MA")) +
  scale_color_manual(values = c("Close" = "steelblue", "30-Day MA" = "red")) +
  labs(
    title = "BTC-USD: Close vs 30-Day Moving Average",
    x = "Date",
    y = "Price (USD)",
    color = "Series"
  )

```
