---
title: "5b-ELO-calculations"
format: html
---

## Introduction

This analysis evaluates player performance in a chess tournament by comparing actual scores
against ELO-expected scores. The dataset (`chess_tournament_players.csv`) contains 64 players
with their pre-ratings, total points scored, and average opponent pre-ratings.

**Approach:**

1. For each player, I compute the expected score per game using the standard ELO
expected-score formula, then multiply by the number of rounds (7) to get a total expected
score.

2. Comparing this to the actual score reveals which players most overperformed or
underperformed relative to expectations.

3. Then rank players by the difference
(actual − expected) to identify the top five overperformers and top five underperformers.

The ELO expected score formula used is:

$$E = \frac{1}{1 + 10^{(R_{\text{opp}} - R_{\text{player}}) / 400}}$$

Source: https://en.wikipedia.org/wiki/Elo_rating_system#Mathematical_details


## Solution

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(knitr)
```

### 1. Load the Data

```{r}
path = "https://raw.githubusercontent.com/qingquan-li/Data607/refs/heads/main/week4/chess_tournament_players.csv"
players <- read.csv(path, stringsAsFactors = FALSE)

# Clean up column names for convenience
colnames(players) <- c("Name", "State", "TotalPoints", "PreRating", "AvgOppRating")

kable(head(players, 10), caption = "First 10 Players in the Dataset")
```

### 2. Calculate Expected Scores

Using the average pre-rating of opponents as a proxy, the expected score per game is:

$$E_{\text{game}} = \frac{1}{1 + 10^{(R_{\text{opp}} - R_{\text{player}}) / 400}}$$

The total expected score over 7 rounds is $E_{\text{total}} = 7 \times E_{\text{game}}$.

```{r}
num_rounds <- 7

players <- players %>%
  mutate(
    # Expected score per game based on ELO formula
    ExpectedPerGame = 1 / (1 + 10^((AvgOppRating - PreRating) / 400)),
    # Total expected score across all rounds
    ExpectedScore = round(num_rounds * ExpectedPerGame, 2),
    # Difference: positive = overperformance, negative = underperformance
    Difference = round(TotalPoints - ExpectedScore, 2)
  )

kable(
  players %>% select(Name, PreRating, AvgOppRating, TotalPoints, ExpectedScore, Difference),
  caption = "All Players with Expected Scores and Differences"
)
```

### 3. Top 5 Overperformers

These players scored the most *above* their ELO-expected score:

```{r}
overperformers <- players %>%
  arrange(desc(Difference)) %>%
  slice_head(n = 5) %>%
  select(Name, State, PreRating, AvgOppRating, TotalPoints, ExpectedScore, Difference)

kable(overperformers, caption = "Top 5 Overperformers (Actual − Expected)")
```

### 4. Top 5 Underperformers

These players scored the most *below* their ELO-expected score:

```{r}
underperformers <- players %>%
  arrange(Difference) %>%
  slice_head(n = 5) %>%
  select(Name, State, PreRating, AvgOppRating, TotalPoints, ExpectedScore, Difference)

kable(underperformers, caption = "Top 5 Underperformers (Actual − Expected)")
```

## Conclusion

Using the standard ELO expected-score formula, we estimated each player's expected
tournament outcome and compared it to their actual performance.

**Findings:**

- The biggest overperformers tended to be lower-rated players who faced relatively strong
  opposition but still managed high point totals.
- The biggest underperformers were often higher-rated players whose point totals fell
  well short of what their rating advantage predicted — possibly due to off-form play,
  tough pairings, or nerves.
